---
title: "01-B-Homework-01"
output: html_document
---

```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# # Define the directory where images generated by knit will be saved
# knitr::opts_chunk$set(fig.path = "images/03-A/")

# Set the web address where R will look for files from this repository
# Do not change this address
# repo_url <- "https://raw.githubusercontent.com/fukamilab/BIO202/master/"
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"
```

Download the data file bleaching.csv into a fresh directory. This data file is derived from a paper on the regional severity of coral bleaching events triggered by high ocean temperatures (McWilliams et al, 2005). 

```{r include = TRUE, eval = FALSE}

```{r, include = TRUE, eval = TRUE}

# List of packages necessary to run this script:
require(librarian, quietly = TRUE)
shelf(tidyverse, cowplot, quiet = TRUE)

# Load Roberts et al. (2022) data:
bleach <- 
  read.csv("https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/data/bleach.csv")

##### Data column descriptions: 
# - Route: name of breeding bird survey route. 
# - Stop: ID for each stop along survey route+
# - Year: year of survey
# - Rich_Grass: grassland bird species richness 
# - Route_factor: name of breeding bird survey route. 
# - Year_Num: survey years re-numbered from 1 (2010) to 7 (2016)
# - Burned: if this stop was burned on or after the given year, 1, else 0
# - Year_Burned: the year that a given stop was burned
# - easting/northing: easting/northing coordinates in UTM 
# - count: number of 30x30m pixels within 400m of each stop. Number of pixels
#          vary because some are masked because they were not 'rangeland' 
#          pixels. See Methods/Data Collection/Tree Cover for details.
# - mean: mean percent tree cover across all 30x30m pixels within 400m of
#         each stop
# - stdDev: standard deviation of percent tree cover across all 30x30m pixels 
#           within 400m of each stop
# - TSF: years-since-fire


```


The first column (SST) is Sea Surface Temperature “anomalies” – that is, the difference between the average sea surface temperature for the Caribbean between Aug and October, and the long run average from 1961 to 1990. Positive values represent years in which the water was warmer than average; negative values represent cool years. The second column in the dataset (BLEACH is the log10 transformed percentage of 1 latitude/longitude cells reporting at least one coral bleaching event in a given year. The final column (MASS) is an indicator marking those years which were considered “Mass Bleaching Events”. 

1)	Load the data in R, and calculate a linear regression model relating LOGBLEACH to SST using any appropriate function. Provide a written interpretation of the results, including an assessment of the assumptions. 

2)	The authors log transformed the bleaching variable “… to generate linear relationships with temperature.” In R, back transform LOGBLEACH to the original percentage scale, and plot against SST. Does a log-transformation make ecological sense, given the type of process and the data? 

3)	At what range of values does the fitted model predict 100% bleaching will occur? (hint: in R you can use the predict function with options newdata and interval=”prediction”. See help(predict.lm) ).

Please do the analyses yourself. Do not use the analysis presented in the paper, although you can refer to the paper for the ecological background. Please provide a script that produces all of your analysis and plots when “sourced” in addition to one file with all answers, in pdf or Microsoft word format. 

McWilliams, J et al (2005) Accelerating impacts of temperature-induced coral bleaching in the Caribbean. Ecology 86(5):2055-2060.





