---
layout: topic
title: "Homework 3"
output: html_document
---

```{r include = FALSE}
# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# # Define the directory where images generated by knit will be saved
# knitr::opts_chunk$set(fig.path = "images/03-A/")

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"

# Suppress warnings and messages
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

## (G)LMMs to the rescue!

One current hypothesis for why some plants are able to invade novel ecosystems is that they “escape” the effect of herbivory by moving to the new environment. Bull Thistle, *Cirsium vulgare*, is a exotic plant introduced to North America from Eurasia. In many parts of the United States it is considered a highly invasive noxious plant.  In Nebraska however, it appears to be present but not reaching high abundances, even after many decades of establishment. There is a native thistle, the Tall Thistle, which in Nebraska flowers simultaneously with Bull Thistle. Many of the native herbivores found on Tall Thistles also attack Bull Thistle, and can consume up to 90% of all viable seeds. Could these native herbivores be suppressing Bull Thistle populations in Nebraska?

Our experiment tested the effects of distance to tall thistle patch on bull thistle seedling establishment density.  We established 4m2 plots of bull thistle seedlings at five distances (0, 7.5, 15, 30 and 60 meters) from a tall thistle patch.  We did not attempt to control the presence or abundance of other patches of tall thistle at the site. At each distance we had both an insecticide (insect herbivory reduced) and control (insect herbivory present) plot.  We sowed a fixed weight of Bull thistle seeds into the plots in April 2005 (538 seeds/ 4m^2) and censused seedling density per 1m2 subplots in September 2005 across all distances for insecticide and control plots.  This experimental design was repeated in 10 sites in Lancaster and southern Saunders Counties.  Additionally, sites are clumped into regions (4 regions) which add a second layer of spatial complexity, assume all sites within a region are closer to each other than in other regions.  Region 1 is 3 sites all at Pioneers park nature center, region 2 is one site at Wilderness Park, region 3 includes 3 sites at Pawnee State Recreational Area and region 4 includes 3 sites at the UNL- ARDC.  A two way ANOVA tells us there is no effect of distance – this is devastating to the enemy release hypothesis and will undo 20 years of my supervisor’s reputation. We did not talk to a statistician before designing the experiment – can you help me?     

## Data description

- Region: the larger spatial grouping of sites.  All sites within the same region are within 1 km of each other.  
- Site: a full set of the experimental treatments
- Distance: distance (m) of bull thistle patch to a tall thistle patch
- Spray: insecticide vs control
- Plot: there are 10 plots at each site, labeled by the spray treatment and distance
- Subplot: Each 4m2 plot is divided into 4-1m2 plots 
		
 
```{r eval = FALSE, message = FALSE, include = TRUE}

# List of packages necessary to run this script:
require(librarian, quietly = TRUE)
shelf(tidyverse,
      MuMIn,
      lme4,
      DHARMa,
      broom.mixed,
      performance,
      quiet = TRUE,
      lib = tempdir())

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/data/"

# Load data
thistles <- read.csv(paste0(repo_url,
                            "C_vulgare_seedling_dataset.csv"))

```

The top row is the insecticide (“i”) plots, bottom row is control plots (“c”). There are four points at each distance for the four subplots. You also need to setup the factors for region, site, plot and subplot thus:

```{r eval = FALSE, message = FALSE, include = TRUE}

thistles$region = factor(thistles$region)
thistles$site = factor(thistles$site)
thistles$plot = factor(paste(thistles$spray, thistles$distance, sep = "."))
thistles$subplot = factor(thistles$subplot)

```