---
layout: topic
title: "Homework 4"
output: html_document
---

```{r include = FALSE}

# This code block sets up the r session when the page is rendered to html
# include = FALSE means that it will not be included in the html document

# Write every code block to the html document 
knitr::opts_chunk$set(echo = TRUE)

# Write the results of every code block to the html document 
knitr::opts_chunk$set(eval = TRUE)

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"

# Suppress warnings and messages
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

## Overview

In this homework assignment, you will analyze a dataset on Forest Bird densities from 56 patches in south eastern Victoria, in Australia. The aim of the study was to relate bird density to 6 habitat covariates: patch area (AREA), distance to nearest patch (DIST), distance to nearest larger patch (LDIST), altitude (ALT), the year the patch was isolated by vegetation clearing (YEAR.ISOL), and a categorical index of stock grazing intensity (GRAZE; 1 = light, 5 = heavy). The data are in the file 'Loyn.txt'. Conduct a full analysis of these data, including data exploration. Be sure to identify non-linear relationships as appropriate, and think carefully about what the appropriate error distribution is for these data. Submit a .Rmd *and* .html file containing a methods section describing your analytical process, and a results section describing the results.

Recall from island biogeography theory that species richness should increase
with AREA and decrease with isolation (DIST, LDIST, and YR.ISOL). Grazing is expected to have a negative effect on forest bird abundance because of the effects on understory vegetation. All species are lumped together in this response.

**NOTE**: You don't need to use an HGAM here, but you can if you feel inclined.


```{r, include = TRUE, eval = TRUE, message = FALSE}

# List of packages necessary to run this script:
require(librarian, quietly = TRUE)
shelf(tidyverse, 
      mgcv, # For checking model convergence
      MuMIn, # for model selection
      gratia, # for ggplot functionality with mgcv
      lib = tempdir(),
      quiet = TRUE)

# Set the web address where R will look for files from this repository
# Do not change this address
repo_url <- "https://raw.githubusercontent.com/LivingLandscapes/Course_EcologicalModeling/master/"

# Load data
loyn <- 
  read.table(paste0(repo_url, "/data/Loyn.txt"), header = TRUE)

# Brief looks at the data:
head(loyn)
pairs(loyn)

```
